\vspace{-2.5em} Excitation and Detection Efficiency Under Light-Sheet Illumination \vspace{-1.0em}

### 2017-06-26 notes/2017-06-21-light-sheet-illumination/report/report.tex
Patrick's June 26 edits.

\begin{document}
\title{\vspace{-2.5em} Excitation and Detection Efficiency Under Scanned Light
  Sheet Illumination \vspace{-2.5em}} \author{Talon Chandler}
\date{\vspace{-1em}\today\vspace{-1em}}
\maketitle
\section{Introduction}
In these notes I will calculate the excitation and detection efficiency of a
single fluorophore under polarized scanned light sheet illumination. I will
start by finding the transverse and longitudinal fields at all positions in a
Gaussian beam using reasonable approximations. Next I will calculate the
excitation efficiency of a single fluorophore as the beam is scanned across the
fluorophore. I will combine these results with the results from previous notes
to write the complete forward model. Finally I will discuss the approximations
we may need to simplify reconstruction.

\section{Transverse and Longitudinal Fields In A Scanned Gaussian Beam}
The complex spatial electric field of a paraxial Gaussian beam propagating along the
$\mh{z}$ axis is \cite{nov}
\begin{align}
  \mb{E}(x,y,z) &= \mb{A}\frac{w_0}{w(z)}\text{exp}\left\{{-\frac{(x^2+y^2)}{w^2(z)}}+i\left[kz - \eta(z) + \frac{k(x^2+y^2)}{2R(z)}\right]\right\}\label{eq:gauss}
\end{align}
where
\begin{alignat}{2}  
  \mb{A} &= \cos\phi_{\text{pol}}\mh{x} + \sin\phi_{\text{pol}}\mh{y}\qquad &&\text{is the input Jones vector},\\
  w_0 &\approx \frac{2n}{k(\text{NA})}\qquad &&\text{is the beam waist radius,}\\
  z_0 &= \frac{kw_0^2}{2} &&\text{is the Rayleigh range,}\\
  k &= \frac{2\pi n}{\lambda}\qquad &&\text{is the wave number,}\\
  w(z) &= w_0\sqrt{1+\frac{z^2}{z_0^2}}\qquad &&\text{is the beam radius,}\\  
  R(z) &= z\left(1+\frac{z_0^2}{z^2}\right)\qquad &&\text{is the wavefront radius,}\\
  \eta(z) &= \text{arctan}\left(\frac{z}{z_0}\right)\qquad &&\text{is the phase correction.}
\end{alignat}
Equation \ref{eq:gauss} uses the paraxial approximation, so it can only be used
for beams with a waist that is significantly larger than the reduced wavelength
($w_0 \gg \lambda/n$). Notice that under the paraxial approximation the beam is uniformly polarized in the transverse plane.

@@

Equation \ref{eq:longit} means that:
\begin{itemize}
\item There is no longitudinal polarization in the $\mh{y}-\mh{z}$ plane
  because $E_z(0,y,z) = 0$.
\item There are longitudinal fields lobes on both sides of the optical axis
  along the transverse polarization direction.
\item The longitudinal fields are $90^{\circ}$ out of phase with the transverse
  fields which means that the total field is elliptically polarized off axis.
\item The longitudinal field strength is proportional to
  $\lambda/w_0^2$---highly focused beams have the strongest longitudinal fields.
\end{itemize}

@@

If the beam is scanned along the $\mh{y}$ axis with velecity $v$ then the time
dependent 3D Jones vector is
\begin{align}
  \mb{A}(x,y,z,t) = \cos\phi_{\text{pol}}\mh{x} + \sin\phi_{\text{pol}}\mh{y} - i\frac{x\cos\phi_{\text{pol}} + (y - vt)\sin\phi_{\text{pol}}}{z_0}\mh{z}\label{eq:scanned_j}
\end{align}
and the time dependent electric field is
\begin{align}
  \mb{E}(x,y,z,t) &= \mb{A}(x,y,z,t)\frac{w_0}{w(z)}\text{exp}\left\{{-\frac{(x^2+(y-vt)^2)}{w^2(z)}}+i\left[kz - \eta(z) + \frac{k(x^2+(y-vt)^2)}{2R(z)}\right]\right\}\label{eq:scanned_e}.
\end{align}
We will use equations \ref{eq:scanned_j} and \ref{eq:scanned_e} to calculate
the excitation efficiency of a single fluorophore. Notice that as the Rayleigh
range $z_0$ increases the longitudinal electric field decreases, so we can ignore the
longitudinal component when the beam is weakly focused.

@@

At first glance equation \ref{eq:ratio} looks bleak---the excitation ratio grows
without bound in the $x$ direction which means that the longitudinal excitation
becomes a larger fraction of the total excitation as we move away from the plane
of the light sheet. Fortunately, we only care about regions of the beam with a
high intensity, so we look at the intensity weighted excitation ratio instead.
\begin{align*}
    \text{Intensity Weighted Excitation Ratio} &= \frac{w_0}{w(z)}e^{-\frac{2x^2}{w^2(z)}}\frac{x^2\cos^2\phi_{\text{pol}} + \frac{1}{4}w^2(z)\sin^2\phi_{\text{pol}}}{z_0^2}\label{eq:intratio} \\
\end{align*}
We can interpret the intensity weighted excitation ratio as the fraction of the
maximum signal (created by a transverse fluorophore at the origin) that we
ignore by ignoring longitudinal excitation. Figure \ref{fig:error} shows that
the intensity weighted excitation ratio is <2\% for the imaging parameters used
in Wu et. al. \cite{wu2013}.

\fig{figures/error.pdf}{1.0}{Intensity weighted excitation ratio as a
  function of position using the parameters in Wu et. al. \cite{wu2013}:
  $w_0 = 1.2\ \mu$m, $z_0 = 9\ \mu$m, $\lambda = 488$ nm, FOV$\approx$ 80$\times$80
  $\mu$m${}^2$. The maximum values are at the edge of the FOV and are <2\%.}{error}

As a rough heuristic, if the fraction of the signal from noise and background is
greater than the intensity weighted excitation ratio we can justifiably ignore
the longitudinal component.

Figure 2 shows the maximum intensity weighted excitation ratio for an
80$\times$80 $\mu$m${}^2$ FOV as a function of $w_0$. The maximum intensity
weighted excitation ratio decreases as $w_0$ increases because wider beams
have smaller longitudinal components.

\fig{figures/max-error.pdf}{0.5}{Maximum intensity weighted excitation ration (see equation \ref{eq:intratio}) for an 80$\times$80 $\mu$m${}^2$ FOV with $\lambda = 488$ nm.}{max-error}

@@

To ignore the longitudinal component, we require that the fraction of the signal
from noise and background is greater than the intensity weighted excitation
ratio (<2\% with current imaging parameters). We'll need to be careful about
longitudinal excitation if we want to use beams that are more strongly focused.

Under the weak focusing approximation the orientation and position of
fluorophores are decoupled. This will allow us to split the reconstruction into
two steps (1) estimate the position of the fluorophores using unpolarized frames
(or the sum of orthogonally polarized frames) with established reconstruction
techniques then (2) estimate the orientation or the fluorophores using polarized
frames and equation \ref{eq:forward}.

>>>>

\begin{document}
\title{\vspace{-2.5em} Excitation and Detection Efficiency Under Light-Sheet
  Illumination \vspace{-1.0em}} \author{Talon Chandler}
\date{\vspace{-1em}June 21, 2017\\ (Updated: June 26, 2017)\vspace{-1em}}
\maketitle
\section{Introduction}
In these notes I will calculate the excitation and detection efficiency of a
single fluorophore under polarized scanned light-sheet illumination. I will
start by finding the transverse and longitudinal fields at all positions in a
Gaussian beam using reasonable approximations. Next I will calculate the
excitation efficiency of a single fluorophore as the beam is scanned across the
fluorophore. I will combine these results with the results from previous notes
to write the complete forward model. Finally I will discuss the approximations
we may need to simplify reconstruction.

\section{Transverse and Longitudinal Fields In A Scanned Gaussian Beam}
The complex spatial electric field of a paraxial Gaussian beam propagating along the
$\mh{z}$ axis is \cite{nov}
\begin{align}
  \mb{E}(x,y,z) &= \mb{A}\frac{w_0}{w(z)}\text{exp}\left\{{-\frac{(x^2+y^2)}{w^2(z)}}+i\left[kz - \eta(z) + \frac{k(x^2+y^2)}{2R(z)}\right]\right\}\label{eq:gauss}
\end{align}
where
\begin{alignat}{2}  
  \mb{A} &= \cos\phi_{\text{pol}}\mh{x} + \sin\phi_{\text{pol}}\mh{y}\qquad &&\text{is the input Jones vector in 3D},\\
  w_0 &\approx \frac{2n}{k(\text{NA})}\qquad &&\text{is the beam waist radius,}\\
  z_0 &= \frac{kw_0^2}{2} &&\text{is the Rayleigh range,}\\
  k &= \frac{2\pi n}{\lambda}\qquad &&\text{is the wave number,}\\
  w(z) &= w_0\sqrt{1+\frac{z^2}{z_0^2}}\qquad &&\text{is the beam radius,}\\  
  R(z) &= z\left(1+\frac{z_0^2}{z^2}\right)\qquad &&\text{is the wavefront radius,}\\
  \eta(z) &= \text{arctan}\left(\frac{z}{z_0}\right)\qquad &&\text{is the phase correction.}
\end{alignat}
Equation \ref{eq:gauss} uses the paraxial approximation, so it can only be used
for beams with a waist that is significantly larger than the reduced wavelength
($w_0 \gg \lambda/n$). Notice that under the paraxial approximation the beam is uniformly polarized in the transverse plane.

@@

Equation \ref{eq:longit} means that:
\begin{itemize}
\item There is no longitudinal polarization in the $\mh{y}-\mh{z}$ plane
  because $E_z(0,y,z) = 0$.
\item There are longitudinal field lobes on both sides of the optical axis
  along the transverse polarization direction.
\item The factor of $i$ means that the longitudinal fields are $90^{\circ}$ out
  of phase with the transverse fields which means that the total field is
  elliptically polarized off axis.
\item The longitudinal field strength is proportional to
  $\lambda/w_0^2$---highly focused beams have the strongest longitudinal fields.
\end{itemize}

@@

If the beam is scanned along the $\mh{y}$ axis with velocity $v$ then the time
dependent 3D Jones vector is
\begin{align}
  \mb{A}(x,y,z,t) = \cos\phi_{\text{pol}}\mh{x} + \sin\phi_{\text{pol}}\mh{y} - i\frac{x\cos\phi_{\text{pol}} + (y - vt)\sin\phi_{\text{pol}}}{z_0}\mh{z}\label{eq:scanned_j}
\end{align}
and the time-dependent electric field is
\begin{align}
  \mb{E}(x,y,z,t) &= \mb{A}(x,y,z,t)\frac{w_0}{w(z)}\text{exp}\left\{{-\frac{(x^2+(y-vt)^2)}{w^2(z)}}+i\left[kz - \eta(z) + \frac{k(x^2+(y-vt)^2)}{2R(z)}\right]\right\}\label{eq:scanned_e}.
\end{align}
We will use equations \ref{eq:scanned_j} and \ref{eq:scanned_e} to calculate
the excitation efficiency of a single fluorophore. Notice that as the Rayleigh
range $z_0$ increases the longitudinal electric field decreases, so we can ignore the
longitudinal component when the beam is weakly focused.

@@

At first glance equation \ref{eq:ratio} looks bleak---the excitation ratio grows
without bound in the $x$ direction which means that the longitudinal excitation
becomes a larger fraction of the total excitation as we move away from the plane
of the light-sheet. Fortunately, we only care about regions of the beam with a
high intensity, so we look at the intensity-weighted excitation ratio instead.
\begin{align*}
    \text{Intensity-Weighted Excitation Ratio} &= \frac{w_0}{w(z)}e^{-\frac{2x^2}{w^2(z)}}\frac{x^2\cos^2\phi_{\text{pol}} + \frac{1}{4}w^2(z)\sin^2\phi_{\text{pol}}}{z_0^2}\label{eq:intratio} \\
\end{align*}
We can interpret the intensity-weighted excitation ratio as the fraction of the
maximum signal (created by a transverse fluorophore at the origin) that we
ignore by ignoring longitudinal excitation. Figure \ref{fig:error} shows that
the intensity weighted excitation ratio is \textless 2\% for the imaging parameters used
in Wu et. al. \cite{wu2013}.

\fig{../figures/error.pdf}{1.0}{Intensity-weighted excitation ratio as a
  function of position using the parameters in Wu et. al. \cite{wu2013}:
  $w_0 = 1.2\ \mu$m, $z_0 = 9\ \mu$m, $\lambda = 488$ nm, FOV$\approx$ 80$\times$80
  $\mu$m${}^2$. The maximum values are at the edge of the FOV and are \textless 2\%.}{error}

As a rough heuristic, if the fraction of the signal from noise and background is
greater than the intensity-weighted excitation ratio we can justifiably ignore
the longitudinal component.

Figure 2 shows the maximum intensity-weighted excitation ratio for an
80$\times$80 $\mu$m${}^2$ FOV as a function of $w_0$. The maximum
intensity-weighted excitation ratio decreases as $w_0$ increases because wider
beams have smaller longitudinal components.

\fig{../figures/max-error.pdf}{0.5}{Maximum intensity-weighted excitation ration (see equation \ref{eq:intratio}) for an 80$\times$80 $\mu$m${}^2$ FOV with $\lambda = 488$ nm.}{max-error}

@@

To ignore the longitudinal component, we require that the fraction of the signal
from noise and background is greater than the intensity weighted excitation
ratio (\textless 2\% with current imaging parameters). We'll need to be careful about
longitudinal excitation if we want to use beams that are more strongly focused.

Under the weak-focusing approximation the orientation and position of
fluorophores are decoupled. This will allow us to split the reconstruction into
two steps (1) estimate the position of the fluorophores using unpolarized frames
(or the sum of orthogonally polarized frames) with established reconstruction
techniques then (2) estimate the orientation or the fluorophores using polarized
frames and equation \ref{eq:forward}.
